Practical Optional Types for Clojure

### 2015-12-31 formal-model.tex
clarify isaprops

We now consider the \isaliteral{} operation, a core part of the dispatch mechanism for multimethods. 
Recalling the \secref{sec:multioverview} examples,
\isaliteral{} is
a subclassing test for classes, otherwise an equality test.
%---we do not model the semantics for vectors
%
The key component of the T-IsA rule (\figref{main:figure:mmsyntax})
is the \isacompareliteral{} 
metafunction, used to calculate the propositions for
\isaliteral{} tests.
%\begin{mathpar}
%  \TIsA{}
%\end{mathpar}

As an example, the expression
\isaapp{\appexp{\classconst{}}{\x{}}}{\Keyword}
has the true and false propositions 
\isacompare{\s{}}{\path{\classpe{}}{\x{}}}{\Value{\Keyword}}{\filterset{\isprop{\Keyword}{\x{}}}{\notprop{\Keyword}{\x{}}}},
meaning that if this expression produces true, \x{} is a keyword, otherwise it is not.

@@

Next, we show how to check
$
{\extendmultiexp {hi_0} {\makekw{en}} {\abs {\x{}} {\Keyword} {\makestr{hello}}}}
$.
%
The expression 
{\extendmultiexp {\e{m}} {\e{v}} {\e{f}}} creates a new multimethod
that extends multimethod \e{m}'s dispatch table, mapping dispatch value
\e{v} to method \e{f}. The T-DefMulti typing rule
checks \e{m} is a multimethod with dispatch function type \t{d},
then calculates the extra information we know based on the current
dispatch value {\thenprop{\proppp{}}}, which is assumed when checking the method
body.
%$$
%    \TDefMethod{}
%$$
Our example checks with \e{m} being of type
$
{\MultiFntype {\ArrowTwo{\x{}}{\Keyword}{\String}}
              {\ArrowOne{\x{}}{\Keyword}{\Keyword}{\filterset{\topprop{}}{\topprop{}}}{\x{}}}}
$
with \objectp{} = {\x{}} (from below the arrow on the right argument of the previous type) and \t{v} = \Value{\makekw{en}}. 
Then {\thenprop{\proppp{}}} = 
{\isprop {\Value{\makekw{en}}}{\x{}}} by
$
\isacompare{}{\x{}}{\Value{\makekw{en}}}
{\filterset {\isprop {\Value{\makekw{en}}}{\x{}}}{\notprop{\Value{\makekw{en}}}{\x{}}}}
$.
Since \t{} = \Keyword{}, we check the method body with
$
\judgement{{\isprop{\Keyword}{\x{}}},{\isprop {\Value{\makekw{en}}}{\x{}}}}
  {\makestr{hello}}
  {\String}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$.
Finally from the interface type \t{m}, we know \thenprop{\prop{}} = \elseprop{\prop{}} = \topprop{},
and \object{} = \emptyobject{}, which also agrees with the method body, above.
Notice the overall type of a \defmethodliteral{} is the same as its first subexpression \e{m}.

>>>>

We now consider the \isaliteral{} operation, a core part of the multimethod dispatch mechanism. 
Recalling the examples in \secref{sec:multioverview},
\isaliteral{} is
a subclassing test for classes, but otherwise is an equality test.
%---we do not model the semantics for vectors
%
The T-IsA rule uses \isacompareliteral{}
(\figref{main:figure:mmsyntax}), a metafunction which produces the propositions for
\isaliteral{} expressions.
%\begin{mathpar}
%  \TIsA{}
%\end{mathpar}

To demonstrate the first \isacompareliteral{} case,
the expression
\isaapp{\appexp{\classconst{}}{\x{}}}{\Keyword}
is true if \x{} is a keyword, otherwise false.
When checked with T-IsA,
the object of the left subexpression \object{} = {\path{\classpe{}}{\x{}}}
(which starts with the {\classpe{}} path element)
and the type of the right subexpression \t{} = {\Value{\Keyword}} (a singleton class type)
together trigger the first \isacompareliteral{} case
\isacompare{\s{}}{\path{\classpe{}}{\x{}}}{\Value{\Keyword}}{\filterset{\isprop{\Keyword}{\x{}}}{\notprop{\Keyword}{\x{}}}},
giving propositions that correspond to our informal description {\filterset{\thenprop{\prop{}}}{\elseprop{\prop{}}}} = {\filterset{\isprop{\Keyword}{\x{}}}{\notprop{\Keyword}{\x{}}}}.

The second \isacompareliteral{} case captures the simple equality mode for non-class singleton types.
For example,
the expression
\isaapp{\x{}}{\makekw{en}} produces true 
when \x{} evaluates to {\makekw{en}}, otherwise it produces false.
Using T-IsA,
it has the propositions {\filterset{\thenprop{\prop{}}}{\elseprop{\prop{}}}} = 
\isacompare{}{\x{}}{\Value{\makekw{en}}}{\filterset {\isprop {\Value{\makekw{en}}}{\x{}}}{\notprop{\Value{\makekw{en}}}{\x{}}}}
since \object{} = {\x{}} and \t{} = {\Value{\makekw{en}}}.
%
The side condition on the second \isacompareliteral{} case ensures we are in equality mode---if \x{} can possibly be a class in 
\isaapp{\x{}}{\Object{}}, \isacompareliteral{} uses its conservative default case,
since if \x{} is a class literal, subclassing mode could be triggered.
%
Capture-avoiding substitution of objects {\replacefor {} {\object{}} {\x{}}} used in this case erases propositions
that would otherwise have \emptyobject{} substituted in for their objects---it
is defined in the appendix.

@@

Next, we show how to check
$
{\extendmultiexp {hi_0} {\makekw{en}} {\abs {\x{}} {\Keyword} {\makestr{hello}}}}
$.
%
The expression 
{\extendmultiexp {\e{m}} {\e{v}} {\e{f}}} creates a new multimethod
that extends multimethod \e{m}'s dispatch table, mapping dispatch value
\e{v} to method \e{f}. The T-DefMulti typing rule
checks \e{m} is a multimethod with dispatch function type \t{d},
then calculates the extra information we know based on the current
dispatch value {\thenprop{\proppp{}}}, which is assumed when checking the method
body.
%$$
%    \TDefMethod{}
%$$
Our example checks with \e{m} being of type
$
{\MultiFntype {\ArrowTwo{\x{}}{\Keyword}{\String}}
              {\ArrowOne{\x{}}{\Keyword}{\Keyword}{\filterset{\topprop{}}{\topprop{}}}{\x{}}}}
$
with \objectp{} = {\x{}} (from below the arrow on the right argument of the previous type) and \t{v} = \Value{\makekw{en}}. 
Then {\thenprop{\proppp{}}} = 
{\isprop {\Value{\makekw{en}}}{\x{}}}
from
$
\isacompare{}{\x{}}{\Value{\makekw{en}}}
{\filterset {\isprop {\Value{\makekw{en}}}{\x{}}}{\notprop{\Value{\makekw{en}}}{\x{}}}}
$
(see \secref{sec:isaformal}).
Since \t{} = \Keyword{}, we check the method body with
$
\judgement{{\isprop{\Keyword}{\x{}}},{\isprop {\Value{\makekw{en}}}{\x{}}}}
  {\makestr{hello}}
  {\String}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$.
Finally from the interface type \t{m}, we know \thenprop{\prop{}} = \elseprop{\prop{}} = \topprop{},
and \object{} = \emptyobject{}, which also agrees with the method body, above.
Notice the overall type of a \defmethodliteral{} is the same as its first subexpression \e{m}.
